---
title: "Brain FANS Project Quality Control within Cell Type"
author: "E Hannon"
output: html_document
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", warning=FALSE, message=FALSE, cache = TRUE)
```
```{r setup, include=FALSE}
library(bigmelon)
library(pheatmap)
source("rmdConfig.run1") ## change the content of this file to to run QC on different set of data
## prior to running this Rmarkdown which summarises the QC output, QC metrics must have been generated
setwd(dataDir) 
load(qcData)
thresBS<-80
passQC<-QCmetrics$intensPASS & QCmetrics$bisulfCon > thresBS & as.character(QCmetrics$predSex) == as.character(QCmetrics$Sex)
load("QCmetrics/CellTypePCA.rdata")

gfile<-openfn.gds(gdsFile, readonly = FALSE)
sampleSheet<-pData(gfile)[passQC,]
studyInd<-sampleSheet$Study
tissueInd<-sampleSheet$Brain.Region
rawbetas<-betas(gfile)[,passQC]
auto.probes<-which(fData(gfile)$chr != "chrX" & fData(gfile)$chr != "chrY")
rawbetas<-rawbetas[auto.probes,]
## exclude rs probes
rawbetas<-rawbetas[-grep("rs", rownames(rawbetas)),]

```

After performing quality control across all samples excluding samples that failed our primary qualtiy control metrics relating to data quality we will perform a secondary QC stage within each cell-type to identify outliers or potentially mislabelled samples. 

First we will produce a heatmap and hierarchical cluster of the samples to gauge how cleanly the cell-types are clustering. The following heatmaps are based on autosomal probes across the top 500 most variable probes in that subset of the data.
 

```{r hclustBefore, fig.width = 15, fig.height = 8, echo = FALSE}

cellTypes<-unique(sampleSheet$Cell.type)
cellCols<-rainbow(length(cellTypes))[as.factor(sampleSheet$Cell.type)]


## first do heatmap to check how things look intially
sample_anno<-sampleSheet[,c("Age","Sex", "Brain.Region", "Indidivual.ID","Cell.type")]
sample_anno$Age<-as.numeric(sample_anno$Age)
rownames(sample_anno)<-sampleSheet$Basename
sigma<-list()

### across everything

sigma[["all"]]<-apply(rawbetas,1, sd)
pheatmap(rawbetas[order(sigma[["all"]], decreasing = TRUE)[1:500],], annotation_col = sample_anno,  show_colnames = FALSE, show_rownames = FALSE, cutree_cols = 4, main = "All")

### split by study, tissue etc.

for(eachStudy in unique(studyInd)){
  sigma[[eachStudy]]<-apply(rawbetas[,which(studyInd == eachStudy)],1, sd)
  pheatmap(rawbetas[order(sigma[[eachStudy]], decreasing = TRUE)[1:500],which(studyInd == eachStudy)], annotation_col = sample_anno[which(studyInd == eachStudy),],  show_colnames = FALSE, show_rownames = FALSE, cutree_cols = 4, main = eachStudy)
}

for(eachBR in unique(tissueInd)){
  sampleKeep<-which(tissueInd == eachBR)
  sigma[[eachBR]]<-apply(rawbetas[,sampleKeep],1, sd)
  pheatmap(rawbetas[order(sigma[[eachBR]], decreasing = TRUE)[1:500],sampleKeep], annotation_col = sample_anno[sampleKeep,],  show_colnames = FALSE, show_rownames = FALSE, cutree_cols = 4, main = eachBR)
}

for(eachStudy in unique(studyInd)){
  for(eachBR in unique(tissueInd)){
    sampleKeep<-which(studyInd == eachStudy & tissueInd == eachBR)
    if(length(sampleKeep) > 2){
      sigma[[paste(eachStudy, eachBR, sep = "_")]]<-apply(rawbetas[,sampleKeep],1, sd)
	
	    pheatmap(rawbetas[order(sigma[[paste(eachStudy, eachBR, sep = "_")]], decreasing = TRUE)[1:1000],sampleKeep], annotation_col = sample_anno[sampleKeep,],  show_colnames = FALSE, show_rownames = FALSE, cutree_cols = 4, main = paste(eachStudy, ":", eachBR))
    }
  }
}


```

Next we will do PCA across all samples and use Mahalanobis distance to ascertain if each sample looks like the rest of it's cell type.

```{r mahalanobisDist, eval = FALSE}


```

Next we will use principal component analysis within each cell-type to identify outlier samples. Within each cell-type we take all the PCs that explain > 1% of the variance. In the histograms below the dashed red lines indicate 2 or 3 SD from the mean.

```{r pcaPlots, fig.width = 15, fig.height = 8, echo = FALSE}

for(study in unique(studyInd)){
	for(each in cellTypes){
		sampleInd<-which(sampleSheet$Cell.type == each & sampleSheet$Study == study)
		if(length(sampleInd) > 5){
		pca<-cellPCA[[study]][[each]]
		par(mfrow = c(1,1))
		varExp<-pca$sdev^2/sum(pca$sdev^2)*100
		plot(1:30, varExp[1:30]*100, type = "b", ylab = "% variance explained", xlab = "Beta PC", lty = 1, main = paste(study, each, sep = " : "))
		abline(h=1, col = "red")
		par(mfrow = c(2,3))
		for(j in which(varExp > 1)){
			pcDat<-pca$x[,j]
			mu<-mean(pcDat, na.rm = TRUE)
			sigma<-sd(pcDat, na.rm = TRUE)
			#x_lim<-range(c(mu-3*sigma, mu+3*sigma, pcDat), na.rm = TRUE)
			hist(pcDat, xlab = paste("Betas:PC", j, sep = ""), breaks = 15, main = paste(signif(varExp[j],3), "% variance explained"), col = "gray", cex.axis = 1.5, cex.lab = 1.5, cex.main = 1.5)
			for(i in 2:3){
				abline(v = mu+i*sigma, col = "red", lty = 2)
				abline(v = mu-i*sigma, col = "red", lty = 2)
			}
		}	
		
		par(mfrow = c(1,3))
		for(j in which(varExp[-1] > 0.01)){
			j<-j+1
			plot(pca$x[,1], pca$x[,j], pch = 16, xlab = "Betas:PC1", ylab = paste("Betas:PC", j, sep = ""), main="")
		}
		}
	}
}

```
