---
title: "Optimising state number"
author: "Jessica Shields"
date: "05/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', results = FALSE)

library(stringr)
library(ggplot2)
library(data.table)
library(RColorBrewer)
library(gtools)
library(grid)
library(gridExtra)
library(prodlim)
library(ComplexHeatmap)

readEmissionTable<-function(tablePath){
  t<-read.table(tablePath, sep='\t', header = TRUE, row.names = 1) %>% 
    as.matrix() %>%
    .[, sort(colnames(.))] %>% # reorder to same col order
  {(.>=0.5)*1} # binarise dataframe (threshold 0.5)
  return(t)
}

euclidean <- function(a, b) sqrt(sum((a - b)^2))


## load arguments
args = commandArgs(trailingOnly=TRUE)
if (length(args)==0){
  args<-"simulateRM"
} 
intproject<-args
type = 'PFC'

source("/lustre/home/jms260/BrainFANS/integrative/chromHMM/config/config.r")

```

## Dataset 'completeness'

A simple measure of completeness is the average number of times a mark is observed across the genome. 

```{r zeros}
files<-list.files(mergeDir, pattern = 'chr\\d+_binary.+') %>%
  mixedsort()

prop<-lapply(files, function(x){
  tmp<-read.delim(paste(mergeDir,x, sep='/'), skip=1)
  apply(tmp, 2, function(x){sum(x)/nrow(tmp)})
}) %>% 
  as.data.frame(col.names = 1:length(.)) %>%
  t()

reshape2::melt(prop) %>% 
  ggplot(aes(Var2, value))+
  geom_boxplot()+
  xlab('Mark')+
  ylab('Abundance')+
  theme_bw()
```

## Optimising state number

### 1.1 Akaike's Information Criterion: a statistical method for optimising model choice

Akaike's Information Criterion (AIC) is a statistical measure which offsets the gain in likelihood of the model against an increasing number of parameters. The lower the AIC, the better the model. However, when considering chromatin states, optimising this criterion invariably results in the model with the higher number of states being selected. The metric should therefore be considered alongside others in order to optimise state number according to both likelihood and feasibility of human understanding.

The decrease in AIC occurs with diminishing improvements, therefore can be used here to indicate a lower bound for the number of states.

```{r cars}
files<-Sys.glob(file.path(paste0(qcDir, '/likelihood.*.txt')))

likeli<-NULL
for (each in files){
  if (file.info(each)$size != 0){ #if file not empty
    tmp<-read.table(each)
    colnames(tmp)<-('logLikelihood')
    tmp$states<-gsub(paste0(qcDir, '/likelihood.'), '',each) %>%
        gsub(".txt", "", .) %>%
        as.numeric()
    likeli<-rbind(likeli, tmp)
  }
}

likeli$params<-(likeli$states*5)+(likeli$states^2) # emission probs + transition probs
likeli$aic<-(2*likeli$params)-(2*likeli$logLikelihood)

ggplot(likeli, aes(states, aic))+
  geom_point()+
  theme_bw()+
  geom_line()
```

## Unique combinations

The below heatmap shows the combinations of marks that appear at least once across all the models.

```{r unique, fig.height=5, fig.width=3}
Efiles<-list.files(modelDir, pattern = 'emissions_..?.txt') %>%
  mixedsort()

uniq<-NULL
for (each in Efiles){
  uniq<-rbind(uniq, readEmissionTable(paste(modelDir, each, sep = '/')) %>%
                as.data.frame())
}

uniq<-unique(uniq) %>%
  as.matrix()
rownames(uniq)<-seq(1, nrow(uniq))
uniq

Heatmap(uniq, col = brewer.pal(9,"YlOrRd"), show_row_dend = FALSE, 
                                         show_column_dend =FALSE,show_heatmap_legend = FALSE, 
                                         cluster_rows = FALSE, cluster_columns = FALSE)
```

## Heatmap of combinations



```{r heatmaps, fig.width = 8, fig.height=13}
#emis<-NULL
#for (x in 1:length(files)){
#  emis[[x]]<-readEmissionTable(paste(modelDir, files[x], sep = '/'))%>%
#    as.data.frame()
#  
  # create list of states matched to the unique combinations
#  tmp<-row.match(matrix(emis[[x]]), uniq)
#  
#  emis[[x]]<-as.matrix(emis[[x]])
##  rownames(emis[[x]])<-tmp
#}


#plots<-NULL
#for (z in 1:(length(emis)-1)){
#  euc<-matrix(data =NA, nrow=nrow(emis[[z]]), ncol=nrow(emis[[z+1]]))
#  for (x in 1:nrow(emis[[z]])){
#    for (y in 1:nrow(emis[[z+1]])){
#      euc[x, y]<-euclidean(emis[[z]][x,], emis[[z+1]][y,])
#    }
#  }
#  euc<-euc/max(euc)
#  
#  rownames(euc)<-rownames(emis[[z]])
#  colnames(euc)<-rownames(emis[[z+1]])
#  
#  euc<-euc[mixedsort(rownames(euc)), mixedsort(colnames(euc))]
#  
#  plots[[z]]<-grid.grabExpr(draw(Heatmap(euc, col = brewer.pal(9,"YlOrRd"), show_row_dend = FALSE, 
#                                         show_column_dend =FALSE,show_heatmap_legend = FALSE, 
#                                         cluster_rows = FALSE, cluster_columns = FALSE)))
#}
#grid.newpage()
#grid.arrange(grobs=plots, ncol=3, layout)
```

## Transition matrix metrics

The transition probability of the model control the way in which the hidden state, in this case the matrices of the models 

```{r transitionModel}
Tfiles<-list.files(modelDir, pattern = 'transitions_..?.txt') %>%
  mixedsort()


emis<-NULL
trans<-NULL
for (x in 1:length(Efiles)){
  emis[[x]]<-readEmissionTable(paste(modelDir, Efiles[x], sep = '/'))%>%
    as.data.frame()
  
  # create list of states matched to the unique combinations
  tmp<-row.match(matrix(emis[[x]]), uniq)
  
  emis[[x]]<-as.matrix(emis[[x]])
  rownames(emis[[x]])<-tmp
  
  trans[[x]]<-read.table(paste(modelDir, Tfiles[x], sep='/'), sep='\t', header = TRUE, row.names = 1) %>% 
    as.matrix()
  
  rownames(trans[[x]])<-tmp
  colnames(trans[[x]])<-tmp
}



tmp<-sapply(trans, function(i){apply(i, 1, function(x){sum(x)-max(x)})}) #%>%
transProb<-sapply(tmp, function(x){sum(x)/length(x)}) %>%
  cbind(sapply(trans, nrow), .)


plot(transProb)

cbind(names(tmp[[1]]), tmp[[1]]) %>%
  plot()

unlist(tmp) %>%
  melt()

```

## Genomic coverage of each state as state number increases


```{r genomCov,fig.width = 8, fig.height=9}
## for the increasing number of states on the whole genome
set.seed(150)
brbg <- c(brewer.pal(12, 'Set3'), brewer.pal(3, 'Dark2')) %>%
  sample()

## create long matrix with order of states 
cov<- unlist(lapply(emis, rownames)) %>%
  matrix(nrow=length(.), ncol = 3)

# read in each of the genomic coverages and add to long_pivoted df
cov[,2]<-unlist(lapply(emis, function(x){
                                rep(paste0('E', nrow(x)), nrow(x))}))

cov[,3]<-unlist(lapply(emis, function(x){read.table(paste0(modelDir, "/", type, "_",nrow(x),"_overlap.txt"), sep='\t', header=TRUE)[c(1:nrow(x)),2]})) 


cov[is.na(cov)]<-0
cov[cov==100]<-0

cov<-as.data.frame(cov) %>%
  setnames(c('State', 'Model', 'genomeCov'))
cov$genomeCov<-as.numeric(levels(cov$genomeCov))[cov$genomeCov]

cov[,2]<-factor(cov[,2], levels = unique(cov[,2]))

set.seed(120)
brbg <- c(brewer.pal(9,"YlOrRd"), brewer.pal(3, 'Dark2')) %>%
  sample()


ggplot(cov, aes(fill = State, x = Model, y = genomeCov))+
  geom_bar(stat = 'identity', position = 'stack', width = 0.7, colour = "black") +
#  scale_fill_manual(values = c24)+
  theme_bw()

```


