---
title: "ChromHMM optimal states: collate bootstrap metrics "
author: "Jessica Shields"
date: "05/10/2022"
output: html_document
##runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results=FALSE, fig.align = "center")

library(ggplot2)
library(gridExtra)
library(magrittr)
library(reshape2)
library(data.table)
library(stringr)
library(ComplexHeatmap)
library(gplots)
library(RColorBrewer)
#library(shiny)

colOrder<-c("h3k27me3", "X5mc", "h3k27ac", "atac", "X5hmc") 

## load arguments
args = commandArgs(trailingOnly=TRUE)
if (length(args)==0){
  args<-"bootstrap"
} 
intproject<-args

source("/lustre/home/jms260/BrainFANS/integrative/chromHMM/config/config.r")

files<-Sys.glob(file.path(paste0(chromDir, '/**/3_model/emissions_10.txt'))) %>%
  sort()


dataDir<-"/lustre/projects/Research_Project-MRC190311/integrative/chromHMM/bootstrap/"
suffix<-"/3_model/emissions_10.txt"
euclidean <- function(a, b) sqrt(sum((a - b)^2))

```

```{r functions}
readEmissionTable<-function(tablePath){
  t<-read.table(tablePath, sep='\t', header = TRUE, row.names = 1) %>% 
    as.matrix() %>%
    .[, colOrder] %>% # reorder to same col order
    {(.>=0.5)*1} # binarise dataframe (threshold 0.5)
  return(t)
}
```

# Optimising state number

## 1. Complete dataset

### 1.1 Akaike's Information Criterion: a statistical methods for optimising model choice

Akaike's Information Criterion (AIC) is a statistical measure which offsets the gain in likelihood of the model against an increasing number of parameters. The lower the AIC, the better the model. However, when considering chromatin states, optimising this criterion invariably results in the model with the higher number of states being selected. The metric should therefore be considered alongside others in order to optimise state number choice to both likelihood and feasibility of human understanding.

The decrease in AIC occurs with diminishing improvements, therefore is used here to indicate a lower bound for the number of states.

```{r aic}
files<-Sys.glob(file.path(paste0(logDir, '/**/likelihood.*.txt')))

likeli<-NULL
for (each in files){
  tmp<-read.table(each)
  colnames(tmp)<-('logLikelihood')
  tmp$states<-substring(each, 87, str_length(each)) %>%
    gsub(".txt", "", .) %>%
    as.numeric()
  likeli<-rbind(likeli, tmp)
}

likeli$params<-(likeli$states*5)+(likeli$states^2) # emission probs + transition probs
likeli$aic<-(2*likeli$params)-(2*likeli$logLikelihood)

ggplot(likeli, aes(states, aic))+
  geom_point()+
  theme_bw()+
  geom_line()

```


## 2. Bootstrapped dataset

This part of the report will look at the 

### 2.1 Akaike's Information Criterion

### 2.2 

Comparing splitwise the proportion of states (combinations of marks) that are conserved between various splits of the data. e.g. if in one half of the split 

Boxplot showing the variation in correlation between each split of the bootstrap across states

```{r splits}
states<-c(5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15)
bootstrap<-c(seq(0, 18, 2))#, seq(22, 40, 2))
corr.all<-matrix(NA, nrow=length(states), ncol=length(bootstrap))

## internal for loop reads in the two emission probabilities and 
for (z in 1:length(states)){
  metric<-NULL
  suffix<-paste0("/3_model/emissions_", states[z], ".txt")
  y=1
  for (each in bootstrap){
    models<-c(paste0(dataDir, each, suffix), paste0(dataDir, each+1, suffix))
    emis<-NULL
    
    for (x in 1:2){
      emis[[x]]<-readEmissionTable(models[x]) %>%
        split(seq(nrow(.))) # convert matrix rows into single list entry
    }
    
    compare<-cbind(emis[[1]] %in% emis[[2]], 
                   emis[[2]] %in% emis[[1]])
    metric[[y]]<-(1-(length(compare)-sum(compare[,1])-sum(compare[,2]))/10)
    y<-y+1
  }
  corr.all[z,]<-metric
}
## matrix where the cols are bootstrap and rows are states
rownames(corr.all)<-paste0('E', seq(states[1],states[length(states)]))

reshape2::melt(corr.all) %>%
  ggplot(aes(Var1, value))+
  geom_violin()+
  stat_summary(fun="mean", 
               geom="point", color="black")+
  xlab("Number of States in Model")+
  theme_bw()




```

Comparing splitwise the euclidean distance between  of states (combinations of marks) that are conserved between various splits of the data. e.g. if in one half of the split 

```{r btstrapEuclidean}
## internal for loop reads in the two emission probabilities and 
euc.all<-matrix(NA, nrow=length(states), ncol=length(bootstrap))
for (z in 1:length(states)){
  metric<-NULL
  suffix<-paste0("/3_model/emissions_", states[z], ".txt")
  k=1
  for (each in bootstrap){
    print(k)
    models<-c(paste0(dataDir, each, suffix), paste0(dataDir, each+1, suffix))
    emis<-NULL
    #print(models)
    
    for (x in 1:2){
      emis[[x]]<-readEmissionTable(models[x])
    }
        
    euc<-matrix(data =NA, nrow=nrow(emis[[1]]), ncol=nrow(emis[[2]]))
    for (x in 1:nrow(emis[[1]])){
      for (y in 1:nrow(emis[[2]])){
        euc[x, y]<-euclidean(emis[[1]][x,], emis[[2]][y,])
      }
    }
    euc<-euc/max(euc)
    
    metric[[k]]<-apply(euc, 2, min) %>%
      mean()
    print(metric)
    k<-k+1
  }
  euc.all[z,]<-metric
}
## matrix where the cols are bootstrap and rows are states
rownames(euc.all)<-paste0('E', seq(states[1],states[length(states)]))

reshape2::melt(euc.all) %>%
  ggplot(aes(Var1, value))+
  geom_violin()+
  stat_summary(fun="mean", 
               geom="point", color="black")+
  xlab("Number of States in Model")+
  theme_bw()
```


# Compare between the first split of the data



## Conserved states

Heatmap showing the 

```{r pressure, echo=FALSE}
states<-c(5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15)
plots<-NULL
noState<-matrix(nrow=length(states), ncol=2)
noState[,1]<-states
for (z in 1:length(states)){
  suffix<-paste0("/3_model/emissions_", states[z], ".txt")
  emis<-NULL
  filePath<-paste0(dataDir, 0, suffix)
  emis[[1]]<-readEmissionTable(filePath) %>%
    split(seq(nrow(.))) # convert matrix rows into single list entry
  for (x in c(1:19,22:41)){
    filePath<-paste0(dataDir, x, suffix)
    #print(filePath)
    emis[[x]]<-readEmissionTable(filePath) %>%
      split(seq(nrow(.))) # convert matrix rows into single list entry
    emis[[1]]<-Reduce(intersect,list(emis[[1]],emis[[x]]))
    }
  #print(length(emis[[1]]))#
  noState[z,2]<-length(emis[[1]])
  consvd<-emis[[1]] %>%
    as.data.frame(., col.names=paste0("E", seq(len=length(.))), row.names = colOrder)%>%
    t()
  
  print(heatmap.2(consvd, scale = "none", Rowv=FALSE, Colv = FALSE, 
             trace="none", col = brewer.pal(9,"YlOrRd"), keysize = 0.1, key=FALSE, 
             margins = c(9,9)))
}
knitr::knit_exit()
```

## Distribution of segment lengths

```{r seglength}
df<-fread("/lustre/projects/Research_Project-MRC190311/integrative/chromHMM/bootstrap/0/3_model/N+_10_segments.bed")

df$V5<-df$V3-df$V2
df$V6<-log(df$V5)

ggplot(df, aes(V4, V6))+
  geom_boxplot()+
  theme_bw()

df2<-fread("/lustre/projects/Research_Project-MRC190311/integrative/chromHMM/bootstrap/1/3_model/N+_10_segments.bed")
df2$V5<-df2$V3-df2$V2
df2$V6<-log(df2$V5)

ggplot(df2, aes(V4, V6))+
  geom_boxplot()+
  theme_bw()
```

## Genome coverage of each state

```{r gencov}

## the genomic coverage of each label 

cov<-read.table("/lustre/projects/Research_Project-MRC190311/integrative/chromHMM/bootstrap/0/3_model/N+_10_overlap.txt", sep='\t', header=TRUE)
cov<-cov[c(1:10),c(1,2)]
colnames(cov)<-c('State', 'split01')
split02<-read.table("/lustre/projects/Research_Project-MRC190311/integrative/chromHMM/bootstrap/1/3_model/N+_10_overlap.txt", sep='\t', header=TRUE) %>%
  .[c(1:10),2]

cov<-cbind(cov, split02)

reshape2::melt(cov) %>%
  ggplot(., aes(fill = State, x = variable, y = value))+
  geom_bar(stat = 'identity', position = 'stack')+
  theme_bw()+
  scale_fill_brewer(palette = 'Set3')
```