---
title: "QC Stage 3: Check Cell Type"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
params:
  configFile: !r commandArgs(trailingOnly=TRUE)[1]
---


```{r setup, include=FALSE}

knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.align='center', echo = TRUE)
options(scipen=5)

library(plotly)
library(Rsubread)
library(scater)
library(GenomicRanges)
library(Matrix)
library(edgeR)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(gridExtra)
library(ggpubr)
library(DESeq2)
library(ggplot2)
library(pheatmap)
library(RColorBrewer)
library(dplyr)
library(cluster)
library(BSgenome.Hsapiens.UCSC.hg38)
library(ggExtra)
library(corrplot)
library(kableExtra)
library(DT)
library(EDASeq)

source(params$configFile)

```

```{r}

## Load data and samples metadata
metadata<-read.table(sampleSheet, header = TRUE, sep = ',', stringsAsFactors = FALSE)
samplesPassed <- read.csv(file.path(qc12stats), sep = ",",stringsAsFactors = FALSE)
samples <- read.csv(file.path(groupAnalysisSamples), sep = ",",stringsAsFactors = FALSE)
samplesPassed <- samplesPassed[samplesPassed$PASSALL == TRUE,]
qcStats <- read.table(qc1stats, header = TRUE, sep = ',', stringsAsFactors = FALSE)
metadata$sequencingBatch <- as.factor(metadata$sequencingBatch)
metadata<- droplevels(metadata)
qcStats <- qcStats[qcStats$sampleID %in% metadata$sampleID,]

genome <- getBSgenome("hg38")
df_ordered <- metadata[order(metadata$fraction), ]
groups <- as.factor(df_ordered$fraction)
colors <- brewer.pal(3, "Set2")
```

```{r}

##Functions

## Load counts in peaks files and match to pheno samples data
loadCounts <- function(counts){
  colnames(counts) <- lapply(colnames(counts),function(x) gsub("*.filt.nodup.bam", "",x))
  counts <- counts[, colSums(is.na(counts)) == 0]
  return(counts)
}

## Create Deseq object
DDSqobj <- function(counts, pheno){
  group <- factor(pheno$fraction)
  metaData <- data.frame(group, row.names = colnames(counts))
  dds <- DESeqDataSetFromMatrix(countData = counts, colData = metaData, design = ~ group)
  atacDDS<- DESeq(dds)
  return(atacDDS)
}

## Plot PCA with first two principal components
pca_single <- function(counts,pheno_sub){
  pca <- prcomp(t(counts))
  tpm.scores = pca$x
  tpm.pca<-pca$sdev^2/sum(pca$sdev^2)
  tpm.scores<-tpm.scores[,which(tpm.pca > 0.01)]
  percentVar = round(100 * (tpm.pca) )
  pheno_1 = pheno_sub[match(rownames(tpm.scores),pheno_sub$sampleCode),] 
  tpm.scores.df <- data.frame(pc1 = tpm.scores[,1],pc2 = tpm.scores[,2],fraction = as.factor(pheno_1$fraction), cohort = as.factor(pheno_1$cohort),sampleCode =as.factor(pheno_1$sampleCode), gender =as.factor(pheno_1$"gender") )

  plot1 <- ggplot(tpm.scores.df, aes(x=pc1, y=pc2, color = fraction, label= sampleCode))+geom_point(size=2)+ theme_bw()+
    labs(x=paste0("PC1: ", percentVar[1], "% variance"),y=paste0("PC2: ", percentVar[2], "% variance")) +theme(legend.position="bottom")
  
  return(plot1)
}

## Plot PCA with first four principal components
pca_multiple <- function(counts,pheno_sub){
  pca <- prcomp(t(counts))
  tpm.scores = pca$x
  tpm.pca<-pca$sdev^2/sum(pca$sdev^2)
  tpm.scores<-tpm.scores[,which(tpm.pca > 0.01)]
  percentVar = round(100 * (tpm.pca) )
  pheno_1 = pheno_sub[match(rownames(tpm.scores),pheno_sub$sampleCode),] 
  tpm.scores.df <- data.frame(pc1 = tpm.scores[,1],pc2 = tpm.scores[,2], pc3 = tpm.scores[,3],pc4 = tpm.scores[,4],fraction = as.factor(pheno_1$fraction), cohort = as.factor(pheno_1$cohort),sampleCode =as.factor(pheno_1$sampleCode), gender =as.factor(pheno_1$"gender") )

  plot1 <- ggplot(tpm.scores.df, aes(x=pc1, y=pc2, color = fraction))+geom_point(size=2)+ theme_bw()+
    labs(x=paste0("PC1: ", percentVar[1], "% variance"),y=paste0("PC2: ", percentVar[2], "% variance")) +theme(legend.position="none")
  
  plot2 <- ggplot(tpm.scores.df, aes(x=pc3, y=pc4, color = fraction)) + geom_point(size=2)+theme_bw() +
    labs(x=paste0("PC3: ", percentVar[3], "% variance"),y=paste0("PC4: ", percentVar[4], "% variance")) +theme(legend.position="right")
  
  plot3 <- ggplot(tpm.scores.df, aes(x=pc1, y=pc4, color = fraction)) + geom_point(size=2)+theme_bw() +
    labs(x=paste0("PC1: ", percentVar[1], "% variance"),y=paste0("PC4: ", percentVar[4], "% variance")) +theme(legend.position="none")
  plot4 <- ggplot(tpm.scores.df, aes(x=pc1, y=pc3, color = fraction)) + geom_point(size=2)+theme_bw() +
    labs(x=paste0("PC1: ", percentVar[1], "% variance"),y=paste0("PC3: ", percentVar[3], "% variance")) +theme(legend.position="none")
  
  figure <- ggarrange(plot1,plot2,plot3,plot4,labels = c("A", "B", "C","D"),ncol = 2, nrow=2)   
  return(figure)
}

## Calculate PCA scores
pca <- function(data){
  pca <- prcomp(t(data))
  tpm.scores = pca$x
  tpm.pca<-pca$sdev^2/sum(pca$sdev^2)
  tpm.scores<-tpm.scores[,which(tpm.pca > 0.01)]
  percentVar = round(100 * (tpm.pca) )
  results <- list(tpm.scores[,c(1:4)], percentVar)
  return(results)
}

## Calculate mean silhouette coefficient
calc.sil <- function(data, y1, name.y1 = "Group1") {
  
  # Check if data is a matrix or data frame and convert if necessary
  if (!is.matrix(data)) data <- as.matrix(data)
  
  # Calculate silhouette scores for y1 (e.g., batch)
  sil_y1 <- silhouette(as.numeric(as.factor(y1)), dist(data))
  mean_sil_y1 <- mean(sil_y1[, 3])  # Mean silhouette width for y1
  
  # Create a list to return results
  result <-list(name = name.y1, silhouette_scores = sil_y1, mean_silhouette = mean_sil_y1)
  
  return(result)
}

mean_sil.1 <- function(silhouttte_coef,coldata){
  sil.df <- as.data.frame(silhouttte_coef$silhouette_scores)
  mean_values <- sil.df %>%
    group_by(cluster) %>%
    summarize(mean_value = mean(sil_width))
  mean.df <- as.data.frame(mean_values)
  batch <- as.factor(coldata$fraction)
  mean.df$cluster <- levels(batch)
  return(mean.df)
}

sil.1 <- function(silhouttte_coef,coldata){
  sil.df <- as.data.frame(silhouttte_coef$silhouette_scores)[,c(1,3)]
  batch <- as.factor(coldata$fraction)
  sil.df$cluster <- batch
  return(sil.df)
}

pca_silscore <- function(counts,pheno_sub){
  pca <- prcomp(t(counts))
  tpm.scores = pca$x
  tpm.pca<-pca$sdev^2/sum(pca$sdev^2)
  tpm.scores<-tpm.scores[,which(tpm.pca > 0.01)]
  percentVar = round(100 * (tpm.pca) )
  pheno_1 = pheno_sub[match(rownames(tpm.scores),pheno_sub$sampleCode),] 
  tpm.scores.df <- data.frame(pc1 = tpm.scores[,1],pc2 = tpm.scores[,2],fraction = as.factor(pheno_1$fraction), cohort = as.factor(pheno_1$cohort),sampleCode =as.factor(pheno_1$sampleCode), gender =as.factor(pheno_1$"gender"), passed = pheno_1$QCS3)

  plot1 <- ggplot(tpm.scores.df, aes(x=pc1, y=pc2, color = fraction, shape = passed, label= sampleCode))+geom_point(size=2)+ theme_bw()+scale_shape_manual(values = c("FALSE" = 4, "TRUE" = 16)) +
    labs(x=paste0("PC1: ", percentVar[1], "% variance"),y=paste0("PC2: ", percentVar[2], "% variance")) +theme(legend.position="bottom")
  
  return(plot1)
}

```

## Overview

This is the third stage of the ATAC-seq QC pipeline. After having check the quality and identity of samples, this third stage checks for the assigned cell type of the samples, as well as to identify samples where sorting might have not fully worked. To do this, we will perform a cell type check using the read counts in peaks called at cell type level.

In total, `r nrow(metadata)` were profiled and `r nrow(samplesPassed)` samples passed stage 1 (data quality) and 2 (identity) check of the QC. Below, the number of samples used to perform peak calling at cell type level is shown. Peak calling was performed using MACS3 in paired-end mode and with 1e-3 cutoff for broad peak calling.

```{r}
nCTStart<-table(metadata$fraction)
nCTMid<-table(samplesPassed$fraction)
ctCounts<-cbind(nCTStart, nCTMid)
colnames(ctCounts)<-c("Profiled", "CT Peak Calling")
kbl(ctCounts) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),font_size = 10)

```

## 1.Raw counts

First, we will explore the raw read counts in peaks. 

```{r}
load(paste0(groupsCounts, "/peakCounts_all_filtbam_",macs_thres_group,"_", filtering, ".filt.narrowPeak.rdata"))

counts <- loadCounts(fc_ctPeaks$count)
pheno <- metadata[metadata$sampleID %in% colnames(counts),]
counts <- counts[,match(pheno$sampleID,colnames(counts))]
colnames(counts) <- pheno$sampleCode
qcStats <- qcStats[qcStats$sampleID %in% pheno$sampleID,]

## Prepare peak ranges table
cnt_table <- fc_ctPeaks$annotation
colnames(cnt_table)=c("Geneid","Chr","Start","End","Strand","Length")
cnt_table$Start <- as.numeric(cnt_table$Start)
cnt_table$End <- as.numeric(cnt_table$End)
cnt_table <- cnt_table[complete.cases(cnt_table[c("Start", "End")]), ]

```

The total number of counts in peaks is `r nrow(counts)` and `r ncol(counts)` samples, `r dim(counts)` is dimension of counts and `r dim(pheno)` the sample sheet. In order to reduce noise, peaks with low counts are filtered out. The number of peaks filtered is shown below.

```{r}
dge <- DGEList(counts=counts,
           group=pheno$fraction)
keep <- filterByExpr(dge$counts, group = pheno$fraction)
dge <- dge[keep, ]
counts.filt <- dge$counts

print(nrow(counts)-nrow(counts.filt))
```

### 1.1 Principal Component Analysis

Although filtering has been applied, our count matrix is still too big, so in order to aid visualisation, we select and plot 100,000 random peaks to check the distribution of raw counts. Principal Component Analysis (PCA) is applied on the raw counts.

```{r fig.height = 6, fig.width = 8} 
N <- 100000
sampled_peaks <- sample(rownames(counts.filt), N)
counts.random <- counts.filt[sampled_peaks,]

ggplotly(pca_single(counts.random, pheno), tooltip = "label")
```

```{r fig.height = 8, fig.width = 8} 
pca_multiple(counts.random, pheno)
```

## 2.Normalised counts in peaks

Normalisation is applied to raw counts in peaks to account for biases or factors that we are not interested in. The expression levels of samples become more comparable between and/or within samples.

### 2.1 Variance Stabilizing Transformation (VST)

VST is first used to transform the counts data. This selects 1000 peaks that are 'representative' of the dataset's dispersion trend, and uses the information from these to perform the transformation. The two first PCAs of the transformed data is shown below. The filtered set of counts in peaks is used here.

```{r}
vsd <- vst(counts.filt, blind = FALSE, nsub = 1000)

ggplotly(pca_single(vsd, pheno))
```

```{r fig.height = 8, fig.width = 8} 

counts_to_plot <- vsd
metadata.p <- data.frame(row.names = colnames(counts_to_plot))
metadata.p$fraction <- pheno$fraction
rownames(metadata.p) <- pheno$sampleID
pca_multiple(counts_to_plot, pheno)
```


### 2.2 FQ-FQ normalisation

An alternative normalisation is double Full-quantile normalisation (from EDASeq package). This double normalisation accounts for two main types of effects on peak-level counts:
- within-lane gene-specific (and possibly lane-specific) effects, e.g., related to gene length or GC-content.
- effects related to between-lane distributional differences, e.g., sequencing depth.

The randomly selected subset of counts in peaks is used here for speed and visualisation purposed. Transformed counts are log + 1 in order to perform PCA and plot them.

```{r}

gr = GRanges(seqnames=cnt_table$Chr, ranges=IRanges(cnt_table$Start, cnt_table$End), strand="*", mcols=data.frame(peakID=cnt_table$Geneid))
gr <- gr[gr$mcols.peakID %in% rownames(counts.random),]
peakSeqs = getSeq(x=genome, gr)

gcContentPeaks = letterFrequency(peakSeqs, "GC",as.prob=TRUE)[,1]
cnt_table <- cnt_table[cnt_table$Geneid %in% rownames(counts.random),]
counts.norm <- counts.random
counts.norm <- counts.norm[rownames(counts.norm) %in% cnt_table$Geneid,]
dataWithin <- withinLaneNormalization(counts.norm,gcContentPeaks, which="full")
dataNorm <- betweenLaneNormalization(dataWithin, which="full")

log_counts <- log(dataNorm+1)
ggplotly(pca_single(log_counts, pheno))
```

```{r fig.height = 8, fig.width = 8} 
pca_multiple(log_counts, pheno)
```

## 3. Cell-type check

The main goal of this third part of the pipeline is to check the cell-type assigned to each sample. The counts in cell-type peaks can be used for this, as we will check the clustering of samples based on this. For this, we have perform differential analysis between the cell-groups of samples to obtain the peaks that differentiates the cell-types. The top ` r diffN` differential peaks are used for the clustering of samples and the identification of potential mislabelled samples.

```{r}
load(paste0(groupsCounts, "/cellType_diffPeaks_edgeR_",macs_thres_group,"_", filtering, ".filt.narrowPeak.rdata"))
res_sorted <- results[order(results$FDR, decreasing = FALSE),]
most_significant_peaks <- rownames(head(res_sorted, 5000))
counts.dif <- counts[most_significant_peaks,]
dds.dif <- DESeqDataSetFromMatrix(countData = counts.dif,
                              colData = pheno,
                              design = ~ fraction)
```

### 3.1 VST transformation (normalisation)

We apply VST on these top differential peaks, as well as PCA to investigate the clustering of samples based on cell-types.

```{r}

vsd <- varianceStabilizingTransformation(dds.dif,blind = FALSE)

ggplotly(pca_single(assay(vsd), pheno))
```

We can also explore multiple PCs.
```{r fig.height = 8, fig.width = 8} 

counts_to_plot <- assay(vsd)
pca_multiple(counts_to_plot, pheno)
```

### 3.2 Clustering of samples

We use hierarchical clustering to cluster samples based on counts in peaks (top `r diffN`) and identify groups of samples based on cell-types, sequencing batch or cohort.

```{r ,fig.height = 20, fig.width = 18}

cm = data.frame(as.matrix(counts_to_plot))
pheno$sequencingBatch <- as.factor(pheno$sequencingBatch)
anno = as.data.frame(pheno[,c("fraction", "sequencingBatch", "cohort", "gender")])
rownames(anno) <- pheno$sampleCode
colnames(cm) = rownames(anno)
colnames(anno)<-c("Cell Fraction", "sequencingBatch", "cohort","gender")
rownames(cm) = NULL
my_palette <- colorRampPalette(c("blue", "white", "red"))(50)
breaks <- seq(-2, 2, length.out = 51)
pheatmap(cm, annotation_col = anno,cluster_cols=TRUE,color = my_palette, cluster_rows=FALSE,scale="row",show_rownames=FALSE,show_colnames = FALSE,fontsize_col = 2, clustering_method = "ward.D2",cutree_cols = 4,breaks = breaks)
```

### 3.3 Sample-to-Sample distance

The distance between samples is an additional method to evaluate the similarity between samples within and between cell-fraction groups.

```{r ,fig.height = 20, fig.width = 18}
cm = data.frame(as.matrix(counts_to_plot))
anno = as.data.frame(pheno[,c("fraction","gender", "sequencingBatch", "cohort")])
rownames(anno) <- pheno$sampleCode
colnames(cm) = rownames(anno)
colnames(anno)<-c(" Cell Fraction","gender", "sequencingBatch", "cohort")
my_palette <- colorRampPalette(c("blue", "white", "red"))(50)
sampleDists <- dist(t(cm))
sampleDistMatrix <- as.matrix(sampleDists)
p <-pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = my_palette ,annotation_col=anno,fontsize_col = 2, fontsize_row = 2)
p
pdf(file.path(paste0(countsDir, "/sampleToSampleDist.pdf")), width = 20, height = 20)
print(p)
dev.off()
```

## 4. Silhouette score

The silhouette coefficient (or silhouette score) is a metric used to evaluate the quality of clustering in a dataset. It provides a measure of how similar each point in one cluster is to points in the same cluster compared to points in other clusters. 

- *High score* (close to 1): The sample is well matched to its own cluster and poorly matched to neighboring clusters.
- *Low score* (close to 0): The sample is on or very close to the boundary between clusters.
- *Negative score* (close to −1): The sample is assigned to the wrong cluster.

We use the VST transform counts in peaks to calculate the silhouette coefficient and identify samples that are not in the right cell-type group or that do not cluster with their group.

```{r}
counts.pca.ruv5 <- pca(counts_to_plot)
counts.silh.norm<- calc.sil(counts.pca.ruv5[[1]],colData(vsd)$fraction, name.y1 = 'Cell Fraction')
norm.sil.fraction <- mean_sil.1(counts.silh.norm, colData(vsd))
after.sil.ruv5.fraction <- sil.1(counts.silh.norm,colData(vsd))
```

### 4.1 Identify failed samples

Samples whose silhouette coefficient is lower than `r silScore` are marked as potential samples with wrong cell fraction assigned or low quality samples.

```{r}
after.sil.ruv5.fraction$sample <- pheno$sampleCode
after.sil.ruv5.fraction$cohort <- pheno$cohort

after.sil.ruv5.list <- split(after.sil.ruv5.fraction,after.sil.ruv5.fraction$cluster)
p<- NULL
for(i in 1:length(cellTypes)){
  p[[i]] <- ggplot(after.sil.ruv5.list[[i]], aes(x = sample, y = sil_width, color = cohort)) + 
  geom_point(size=3)+theme_bw()+geom_hline(yintercept =silScore,color = "red", linewidth = 1)+geom_hline(yintercept =0.25,color = "black", linetype = "dashed", linewidth = 1)+geom_hline(yintercept =-0.25,color = "black", linetype = "dashed", linewidth = 1)+
  labs(x = 'Sample', y = 'Silhouette Coefficient', name = 'Batch', title=names(after.sil.ruv5.list)[i]) +theme(legend.title = element_blank(),
        legend.position = 'bottom',axis.text.x = element_text(angle =90, size=7))

}
```

```{r fig.height = 5, fig.width = 9} 
ggplotly(p[[1]])
```

```{r fig.height = 5, fig.width = 9} 
ggplotly(p[[2]])
```

```{r fig.height = 5, fig.width = 9} 
ggplotly(p[[3]])
```

```{r fig.height = 5, fig.width = 9} 
ggplotly(p[[4]])
```

### 4.2 Mark failed samples

We use the first two PCAs to cluster samples based on their assigned cell-types and those that have failed based on the silhouette score threshold are marked with a X.

```{r fig.height = 6, fig.width = 8} 

pheno$QCS3 <- ifelse(after.sil.ruv5.fraction$sil_width < silScore, "FALSE", "TRUE")

ggplotly(pca_silscore(counts_to_plot,pheno),tooltip="label")
```

## Summary

Below, samples that fail at this third QC stage are shown.

```{r, echo = FALSE}
wrongCT <- pheno[pheno$QCS3==FALSE,]

DT::datatable(wrongCT)
```

```{r}
## Collate results from previous Stage 1 and 2 in a single table
samplesPassed <- samplesPassed[samplesPassed$sampleID %in% pheno$sampleID,]
QCPASSALL<-cbind(samplesPassed, "QCS3" = pheno$QCS3)
QCPASSALL$PASSALL[QCPASSALL$QCS2==TRUE & QCPASSALL$QCS1 == TRUE & QCPASSALL$QCS3 == "FALSE"]<- FALSE
QCPASSALL <- QCPASSALL[,c(1:16,18,17)]

write.csv(QCPASSALL, file = file.path(qc3pass),row.names = FALSE)


# generate a txt file of passed samples
write.table(QCPASSALL$sampleID, file = file.path(qc3passList), 
            row.names = FALSE, col.names = FALSE, quote = FALSE)
```

